---
title: "cmp713-project"
author: Hazem Alabiad & Betül Bayrak
date: 25/01/2021
output: rmarkdown::html_vignette
---

# Introduction
Tipping someone is a common thing in our community, especially if this person is a taxi driver, waiter, or delivery officer; The tip amount varies depending on many conditions; In the case of a taxi driver, it might depend on the passenger count, trip distance, trip hour, et cetera.
In this project, we aim to study and predict the tip percentage of the total amount paid to taxi drivers. We used the data provided by NYC Taxi & Limousine Commission (TLC) from this [link](https://www1.nyc.gov/site/tlc/about/tlc-trip-record-data.page).
First, we explore our data to know what our data looks like, find potential outliers, obtain general and basic statistics, for instance, observations count, variables count, NAs' count, et cetera.
Second, we apply what is called data pre-processing and data engineering that include removing unnecessary or NAs' rows, columns, convert the data type to a more suitable one and introducing new variables to get our data ready to be fed our model that is going to predict the tip amount given other variables.
Third, we build our models that utilize the pre-processed, clean data to predict a variable "tip percentage in our case".
Lastly, we measure the performance or accuracy of our model using many approaches and mathematical formulas and plot the predictions vs. actual tip percentage to decide on which model did the best.

# Loading Libraries
```{r}
library(randomForest)
library(ggmap)
library(h2o)
library(gridExtra)
library(dplyr)
library(ggplot2)
library(nnet)
library(hrbrthemes)
library(rpart.plot)
library(mlbench)
library(e1071)
library(rpart)
library(DMwR2)
library(caret)
library(Metrics)
library(fpc)
library(Hmisc)
library(tibble)
library(knitr)
library(MLmetrics)
library(earth) # fit MARS models
```
# Custom Functions
```{r}
`%notin%` <- Negate(`%in%`)

pretty_df_print <- function (df) kable_styling(kable(df))

top_least_freq_and_neg_vals <- function (col) {
  t <- table(col) %>% as.data.frame() %>% arrange(desc(Freq))
  non_pos <- as.data.frame(col)
  non_pos <- non_pos[non_pos <=0, ]
  print("Top frequent values:")
  print.data.frame(head(t))
  print("Least frequent values:")
  print.data.frame(tail(t))
  print("Negative values:")
  print.data.frame(non_pos)
  print("/////////////////////////////////////////////////////")
}

na_count_df <- function (df) {
  return(as.data.frame(t(data.frame(sapply(df, function(col) sum(is.na(col)))))))
}

iqr_outliers_min_max <- function(df) {
  iqr_df <- as.data.frame( t(data.frame(df %>% sapply(function (x) {quantile(x, c(0.25, 0.75))}))) ) %>% cbind((data.frame(df %>% sapply(IQR))))
  colnames(iqr_df) <- c("Q1", "Q3", "IQR")
  iqr_df$minVal <- iqr_df$Q1 - 1.5*iqr_df$IQR
  iqr_df$maxVal <- iqr_df$Q3 + 1.5*iqr_df$IQR
  print.data.frame(iqr_df %>% select("minVal", "maxVal"))
}

iqr_outliers_finder <- function(x){
  Q1 <- quantile(x, 0.25)
  Q3 <- quantile(x, 0.75)
  IQR <- Q3 - Q1
  Vl <- Q1 - 1.5 * IQR
  Vr <- Q3 + 1.5 * IQR
  return (x[which(x < Vl | x > Vr)])
}

get_freq_df_from_vector <- function (vec, col.name) {
  freq_df <- as.data.frame(table(vec))
  colnames(freq_df)[1] <- col.name
  freq_df <- freq_df[order(freq_df[,1], decreasing = T),]
  if(nrow(freq_df) <= 10) {
    print.data.frame(freq_df)}
  else {
    cat("Highest values :\n")
    print.data.frame(head(freq_df))
    cat("Lowest values :\n")
    print.data.frame(tail(freq_df))
  }
}

dbscan_outliers_finder <- function(data, ...) {
  require(fpc, quietly=TRUE)
  cl <- dbscan(data, ...)
  posOuts <- which(cl$cluster == 0)
  list(positions = posOuts,
       outliers = data[posOuts,],
       dbscanResults = cl)
}
```
# Data loading
```{r}
st <- proc.time()
green <- read.csv("data/green_tripdata_2020-06.csv")
green <- green %>% rbind(read.csv("data/green_tripdata_2020-01.csv"))
print(proc.time() - st)
cat('There are [', nrow(green), '] observations, and [',ncol(green),'] variables')
```


# Data exploratory & Preprocessing
## Dataframe head
```{r}
head(green)
```
## Dataframe tail
```{r}
tail(green)
```
## Brief info on the data
```{r}
str(green)
```
## Unnecessary Variables Deletion
```{r}
num.vars <- ncol(green)
green <- green %>% subset(select = -c( VendorID, store_and_fwd_flag))
cat("[", num.vars - ncol(green) , "] variables were delete")
```
## Summary of the data
```{r}
describe(green)
```
## Count NAs' in each column
```{r}
na_count_df(green)
```
## Remove NAs'
As we can see from the output above:
`passenger_count`, `payment_type`, `trip_type`, `congestion_surcharge` have ~ (24 K) NAs'
`fare_amount`, `extra`, `mta_tax`,  `tip_amount`, `tolls_amount`, `improvement_surcharge`, `total_amount` have negative values.
Negative values: we believe have been entered mistakenly as negative values so, we need to fix this problem by obtaining the absolute values.
NAs': We need to drop rows with NAs as they critical in the classification.
As well as, we need to remove `ehail_fee` variable as it's all NAs'.
For `tip_amount` we realized that there are many `0` values, we need to get rid of them.
```{r}
green <- green %>% within(rm(ehail_fee))
num.rows <- nrow(green)
green <- green[!is.na(green$trip_type) & green$trip_distance != 0,]
cat("[", num.rows - nrow(green), "] observations were deleted out of [", num.rows, "]",
    "which means ~ (", round((num.rows - nrow(green))/num.rows, digits = 2), ")")
```
```{r}
as.data.frame(t(data.frame(sapply(green, function(col) sum(is.na(col))))))
```
## Check top & least frequent and negative values
```{r}
describe(green)
```
## Handling `0` values
`passenger_count` variable has some `0` which must be an input error. We can fix this issue by replacing `0` with the rounded mean of all records.
`total_amount` variable has `0` which cannot be handled nor replaced. So, we go with deleting them.
```{r}
avg <- as.integer(round(mean(green$passenger_count), digits = 0))
green$passenger_count <- ifelse(green$passenger_count != 0, green$passenger_count, avg)
green <- green[green$total_amount > 0,]
```
```{r}
str(green)
```
## Convert the negative values to positive by obtaining `abs`
```{r}
green$fare_amount <- abs(green$fare_amount)
green$extra <- abs(green$extra)
green$mta_tax <- abs(green$mta_tax)
green$tip_amount <- abs(green$tip_amount)
green$tolls_amount <- abs(green$tolls_amount)
green$improvement_surcharge <- abs(green$improvement_surcharge)
green$total_amount <- abs(green$total_amount)
```
## Convert categorical variables to factor type
```{r}
green$RatecodeID <- factor(green$RatecodeID)
green$PULocationID <- factor(green$PULocationID)
green$DOLocationID <- factor(green$DOLocationID)
green$payment_type <- factor(green$payment_type)
green$trip_type <- factor(green$trip_type)
```
```{r}
str(green)
```
## Detecting outliers
### Using IQR
After we calculate the outliers using IQR, we inspect each variable to double check and remove manually what we, as exports, believe that it is an outlier.
```{r}
iqr_outliers_min_max(green %>% select_if(is.numeric))
```
```{r}
system.time(iqr.outliers <- green %>% select_if(is.numeric) %>% lapply(iqr_outliers_finder))
```
Now, let us inspect variable values
### passenger_count
```{r}
get_freq_df_from_vector(iqr.outliers$passenger_count, "passenger_count")
```

We can remove observations with `passenger_count > 6` as it is not logical to fit 6 or more people in a taxi.
```{r}
num.obs <- nrow(green)
green <- green[green$passenger_count <= 6, ]
cat("[", num.obs-nrow(green), "] Observation have been deleted! Which is ~ (", round((num.obs-nrow(green))/num.obs, digits = 2), ")")
```
### trip_distance
```{r}
get_freq_df_from_vector(iqr.outliers$trip_distance, "trip_distance")
```
1. Drop the row with `trip_distance == 134121.5`
2. Take the absolute value of the negative ones
```{r}
green$trip_distance <- abs(green$trip_distance)
green <- green[green$trip_distance < 150, ]

num.obs <- nrow(green)
cat("[", num.obs-nrow(green), "] Observation have been deleted! Which is ~ (", round((num.obs-nrow(green))/num.obs, digits = 3), ")")
```
### `fare_amount`
```{r}
get_freq_df_from_vector(iqr.outliers$fare_amount, "fare_amount")
```
`753` is an outlier as the trip distance is only `7.49` => drop
`335.5` is an outlier as the trip distance is only `6.97` => drop
Others seem to be accepted
```{r}
green <- green[green$fare_amount %notin% c(753, 335.5),]

num.obs <- nrow(green)
cat("[", num.obs-nrow(green), "] Observation have been deleted! Which is ~ (", round((num.obs-nrow(green))/num.obs, digits = 3), ")")
```
### `extra`
```{r}
get_freq_df_from_vector(iqr.outliers$extra, "extra")
```
### `mta_tax`
```{r}
get_freq_df_from_vector(iqr.outliers$mta_tax, "mta_tax")
```
### `tip_amount`
```{r}
get_freq_df_from_vector(iqr.outliers$tip_amount, "tip_amount")
```
We remove observations with `tip_amount > 0.7 of the `total_amount`
```{r}
green <- green[(green$tip_amount/green$total_amount) <= 0.7,]

num.obs <- nrow(green)
cat("[", num.obs-nrow(green), "] Observation have been deleted! Which is ~ (", round((num.obs-nrow(green))/num.obs, digits = 3), ")")
```
### `tolls_amount`
```{r}
get_freq_df_from_vector(iqr.outliers$tolls_amount, "tolls_amount")
```

`96.12`, `48.88`, `35.0` and all values with percentage 0.9 or higher of the `total_amount` are outliers to be removed.
```{r}
green <- green[green$tolls_amount/green$total_amount < 0.5, ]

num.obs <- nrow(green)
cat("[", num.obs-nrow(green), "] Observation have been deleted! Which is ~ (", round((num.obs-nrow(green))/num.obs, digits = 3), ")")
```
### `improvement_surcharge`
```{r}
get_freq_df_from_vector(iqr.outliers$improvement_surcharge, "improvement_surcharge")
```
### `total_amount`
```{r}
get_freq_df_from_vector(iqr.outliers$total_amount, "total_amount")
```
### `congestion_surcharge`
```{r}
get_freq_df_from_vector(iqr.outliers$congestion_surcharge, "congestion_surcharge")
```
## Save valid data "after dropping outliers"
```{r}
system.time(saveRDS(green, file = "data/valid_data.rds"))
```
## Load valid data
```{r}
system.time(green <- readRDS("data/valid_data.rds"))
```

## Density distribution of  `trip_distance`
```{r}
ggplot(green, aes(x=trip_distance)) +
        geom_histogram(aes(y=..density..), binwidth=.1, colour="black", fill="white") +
        geom_density(alpha=.2, colour="blue", fill="#000066")+  xlim(0, 15)
```

# Classification
## Add `mean` and `median` of `trip_distance` grouped `hour` of pickup time
```{r}
st <- proc.time()
green$pickup_hour <- as.integer(format(strptime(green$lpep_pickup_datetime, "%Y-%m-%d %H:%M:%S"),"%H"))
hourly_trip_distance <- data.frame( green %>%
                                   group_by(pickup_hour) %>%
                                   summarise(mean_trip_dist = mean(trip_distance),
                                             median_trip_dist = median(trip_distance)) %>% ungroup())
head(hourly_trip_distance)
print(proc.time() - st)
```
```{r}
# Mean trip distance plot
m.trip.dist.plt <- ggplot(hourly_trip_distance, aes(x=pickup_hour, y=mean_trip_dist)) + geom_bar(stat = "identity")
# Median trip distance plot
M.trip.dist.plt <- ggplot(hourly_trip_distance, aes(x=pickup_hour, y=median_trip_dist)) + geom_bar(stat = "identity")
grid.arrange(m.trip.dist.plt, M.trip.dist.plt, ncol=1, nrow =2)
```
From above, we conclude that the longest trips are at `5` & `6` in the morning. In rest of the day's hours trips are somehow close in terms of distance traveled.
//////////////////////////////////////////////////////////////////////////

## `tip_percentage` is the tip percentage based on `total_amount` 
```{r}
st <- proc.time()
green$tip_percentage <- ifelse(green$tip_amount==0.0 | green$total_amount==0.0 , 0.0,
                        round(green$tip_amount/green$total_amount,3))
# Remove all zero percentage as they are not going to help us in classification
num.obs <- nrow(green)
green <- green[green$tip_percentage > 0,]
cat("[", num.obs-nrow(green), "] Observation have been deleted! Which is ~ (", round((num.obs-nrow(green))/num.obs, digits = 3), ")")
print(proc.time() - st)

ggplot(green, aes(x=total_amount, y=tip_amount)) +
    geom_point(
        color="black",
        fill="#69b3a2",
        shape=22,
        alpha=0.3,
        size=3,
        stroke =2
        ) +
    theme_ipsum()
```
`geom_point` is useful when we want to compare two continuous variables.
//////////////////////////////////////////////////////////////////////////

```{r}
cat("Average tip percentage of the total amount ~ (",
    round((sum(green$tip_amount)/sum(green$total_amount)),4)*100 ," %)")
```
## Histogram of `tip_percentage`
```{r}
hist(green$tip_percentage)
```
## Classification using Decision Tree
### Split data into training & testing
```{r}
sample.size <- floor(0.75*nrow(green))
s <- sample(seq_len(nrow(green)), sample.size)
numeic.cols <- green %>% select_if(is.numeric)
train.set <- numeic.cols[s,]
test.set <- numeic.cols[-s, ]
```
```{r}
dt.model <- rpartXse(tip_percentage ~ ., train.set, se = 0.5)
dt.predicted <- round(predict(dt.model, test.set), digits = 3)
saveRDS(file = "data/dt_model.rds", object = dt.model)
saveRDS(file = "data/dt_pred.rds", object = dt.predicted)
head(dt.predicted)
print(proc.time() - st, paste("\n"))
```
## Load DT Model
```{r}
dt.model <- readRDS("data/dt_model.rds")
dt.predicted <- readRDS("data/dt_pred.rds")
```

## Predicted vs original tip percentage using `Random Forest Tree`
```{r}
dt.perf.matrix <- as.data.frame(test.set$tip_percentage) %>% cbind(dt.predicted)
colnames(dt.perf.matrix) <- c("actual", "pred")
dt.perf.matrix["error"] <- dt.perf.matrix["actual"]-dt.perf.matrix["pred"]
dt.perf.matrix["error/actual"] <- abs(dt.perf.matrix["error"]/dt.perf.matrix["actual"])
head(dt.perf.matrix)
```
## Measuring performance using Mean Absolute Percentage Error (MAPE)
```{r}
dt.mape <- mean(dt.perf.matrix$`error/actual`)
cat("Error percentage: (", round(dt.mape, 4), ") \nSuccess percentage: (", round(100-dt.mape, 4),")")
```
```{r}
dt.mse <- mse(test.set$tip_percentage, dt.predicted)
dt.mae <- mae(test.set$tip_percentage, dt.predicted)
dt.rmse <- rmse(test.set$tip_percentage, dt.predicted)
dt.r2 <- R2(test.set$tip_percentage, dt.predicted, form = "traditional")

cat(" MAE:", dt.mae, "\n", "MSE:", dt.mse, "\n",
    "RMSE:", dt.rmse, "\n", "R-squared:", dt.r2)
```
As we can see from above, if we accept only 3 digits after the point, which may cause some lose in data thus error, we obtained a very good success rate using RT.
//////////////////////////////////////////////////////////////////////////
## Using Support Vector Machine (SVM)
```{r}
st <- proc.time()
svm.model <- svm(tip_percentage ~ ., train.set)
svm.predicted <- round(predict(svm.model, test.set), digits = 3)
saveRDS(object = svm.model, file = "data/svm_model.rds")
saveRDS(object = svm.predicted, file = "data/svm_pred.rds")
print(proc.time() - st)
```
## Load SVM Model
```{r}
svm.model <- readRDS("data/svm_model.rds")
svm.predicted <- readRDS("data/svm_pred.rds")
```

## Measuring performance using Mean Absolute Percentage Error (MAPE)
```{r}
svm.perf.matrix <- as.data.frame(test.set$tip_percentage) %>% cbind(svm.predicted)
colnames(svm.perf.matrix) <- c("actual", "pred")
svm.perf.matrix["error"] <- svm.perf.matrix["actual"]-svm.perf.matrix["pred"]
svm.perf.matrix["error/actual"] <- abs(svm.perf.matrix["error"]/svm.perf.matrix["actual"])
head(svm.perf.matrix)
```
```{r}
svm.mape <- mean(svm.perf.matrix$`error/actual`)
cat("Error percentage: (", round(svm.mape, 4), ") \nSuccess percentage: (", round(100-svm.mape, 4),")")
```

```{r}
svm.mse <- mse(test.set$tip_percentage, svm.predicted)
svm.mae <- mae(test.set$tip_percentage, svm.predicted)
svm.rmse <- rmse(test.set$tip_percentage, svm.predicted)
svm.r2 <- R2(test.set$tip_percentage, svm.predicted, form = "traditional")
 
cat(" MAE:", svm.mae, "\n", "MSE:", svm.mse, "\n",
    "RMSE:", svm.rmse, "\n", "R-squared:", svm.r2)
```
## Using Neural Network
```{r}
st <- proc.time()
nn.model <- nnet(tip_percentage ~ ., train.set,
           linout=TRUE,
           trace=FALSE,
           size=6,
           decay=0.01,
           maxit=2000)
nn.pred <- predict(nn.model, test.set)
saveRDS(nn.model, "data/nn_model.rds")
saveRDS(nn.pred, "data/nn_pred.rds")
print(proc.time() - st)
```
## Load NN model
```{r}
nn.model <- readRDS("data/nn_model.rds")
nn.pred <- readRDS("data/nn_pred.rds")
```
```{r}
plot(test.set$tip_percentage, nn.pred)
abline(0, 1)
```
From the plot above, we realize that when the `tip_percentage` is larger the errors, or the residuals become larger.
///////////////////////////////////////////////////////////////////////

## Starting `H2O` Scalable platform that parallelize many machine learning algorithms
```{r}
system.time(h2oInstance <- h2o.init()) # start H2O instance locally
```

## Using `H2O` build a deep neural network with the following parameters
```{r}
st <- proc.time()
trH <- as.h2o(train.set,"trH")
tsH <- as.h2o(test.set,"tsH")
mdl <- h2o.deeplearning(x=1:11, y=12, training_frame=trH,hidden = c(100, 100, 100, 100, 100, 100, 100), epochs = 1000)
preds <- as.vector(h2o.predict(mdl,tsH))
print(proc.time() - st)
```

```{r}
mean(abs(preds - as.vector(tsH$tip_percentage)))
```

```{r}
plot(as.vector(tsH$tip_percentage), preds)
abline(0, 1)
```

```{r}
plot(as.vector(tsH$tip_percentage), preds)
points(as.vector(tsH$tip_percentage), preds, col = "red")
abline(0, 1)
```

```{r}
h2o.shutdown(prompt = F);
```



```{r}
mars1 <- earth(
tip_percentage ~ .,
data = train.set)
```
```{r}
print(mars1)
```

```{r}
summary(mars1) %>% .$coefficients #%>% head(10)
```

```{r}
plot(mars1, which = 1)
```


```{r}
optimal_tree <- rpart(
formula = tip_percentage ~ .,
data = train.set,
method = "anova",
control = list(minsplit = 11, maxdepth = 8, cp = 0.01)
)
plotcp(optimal_tree)
```

# Conclusion
We conclude that data is the chief factor that levels up the model's accuracy so, we first need to be aware of the data and understand it as much as possible. After that, we build the regression model using many available ones such as SVM, Decision Tree, and DNN to predict the tip percentage of the total paid amount given other variables. All the used models gave us similar accuracies of ~ 0.99 that indicates successful training using the provided data.
